A Review of Lossless Data Compression Algorithms

### 2018-12-08 paper.tex
Updates

\usepackage{graphicx}
\usepackage{amssymb, amsmath, amsthm, amsfonts}
\usepackage{thmtools}
\usepackage{tikz}
\usepackage[pdfusetitle]{hyperref}
\usepackage{cleveref}
\usepackage{apacite}

@@

\title{A Review of Lossless Data Compression Methods}
\author{5251 Student}

@@

In the age of ever-expanding quantities of digital data, the ability
to store this data efficiently is more important than ever. Throughout
the history of the field of computer science, algorithms of varying
complexities and efficacies have been created in order to reduce the
amount of space required to store arbitrary strings of binary
data. This wide variety of methods are optimized for different
purposes: some algorithms attempt to reduce the amount of space
required to store raster images; some try to optimize audio storage;
some can optimize genetic data to an extremely high degree; and still
more attempt to work well for general-purpose use. Within the scope of
this paper, I will attempt to give an overview of a handful of popular
compression schemes, including general purpose in addition to
domain-specific methods, as well as a theoretical overview of the
limits and potential benefits of compression in general. This paper
will only cover lossless data encoding: that is, alogrithms which can
reproduce a perfect copy of the input data from its output. Lossy
compression schemes exist which can achieve incredible compression
ratios, but their methods typically include psychological ``tricks''
which allows the algorithm to discard data which is less noticable to
human senses. As such, these methods are outside the scope of this
paper.

@@

Data compression has many benefits as quickly reviewed above; however,
many limitations also exist. Note that for simplicity, we will limit
our discussion of compression schemes to binary schemes, as these are
the most wide spread in applications in the real world (i.e.,
compression of computer files.) The first limitation we shall discuss
is the impossibility of a universally applicable, or ``perfect,''
compression scheme.

@@

The Lempel-Ziv compression algorithm, commonly referred to as LZ77 or
LZ78, or even simply LZ for short, is one of the most widespread
compression schemes in use, especially when including its large family
of derivative methods. It was first introduced in
1977. \cite{LempelZiv} Algorithms based on the Lempel-Ziv algorithms
are relatively efficient and work well with a wide variety of inputs.

The algorithm is based on a dictionary-style approach; that is to say,
it creates a dictionary of codewords within the input data as the
algorithm progresses. Lempel-Ziv uses a ``sliding-window'' technique
to construct its dictionary. For each set of bytes it attempts to
compress, it inspects its sliding window, which extends some set
amount behind the bytes under consideration. In this way, the
algorithm builds a prefix code as it proceeds through the file. A
``prefix'' can be one or more characters long. For each point in the
input data, the algorithm consults the prefix table of prefixes
contained in the sliding window, which is a set amount of data prior
to the current point. If the current prefix is found in the prefix
table, the algorithm outputs a data triplet indicating how far back in
the window the prefix occurred, the length of the prefix, and finally
the character which comes after the prefix. This means that the very
first characters in the input data will be encoded as 1-character
prefixes. (For our very first characters in the input data, since
there can't be any prefixes in the sliding window yet, the encoded
form will consist of zero-length prefixes followed by the input
character literals.)

@@

LZW is able to achieve this speed by using a single symbol table for
the entire input, adaptively building up as the algorithm processes
it. Instead of using a sliding window, LZW instead begins by
initializing a string table with every single-character prefix. As it
parses the input, it uses a greedy algorithm to match strings from the
input to the strings already seen in the table. It does this by
checking if some prefix in the input has already been seen in the
table; if so, it appends the immediately next character from the input
to the prefix and queries the string table again. If, on the other
hand, the appended prefix is \emph{not} found, the algorithm outputs
the code of the prefix \emph{without} the next character appended (the
code for which is retrieved from the string table) and then adding the
appended prefix to the string table. The prefix is set to the last
character of the prefix and the process repeats.

@@

Another algorithm based on the Lempel-Ziv foundation is the DEFLATE
format, codified in RFC 1951. \cite{RFC1951} DEFLATE is not
``standardized'' by any recognized governing body, though it is very
widely used and has become a de facto standard by virtue of its use in
the \texttt{gzip} compression utility, as well as in the ubiquitous
\texttt{zlib} library. RFC 1951 specifies the technical minutia of the
implementation of the file format, as well as its algorithm itself. As
such, it is not useful to describe the full file format within this
paper, only the algorithm will be outlined.

In general, compression proceeds as follows: for a given block of data
withint the input, the algorithm constructs a Huffman prefix coding
for the input alphabet. Information about the prefix coding is then
stored within the output block header, which contains metadata about the
compressed output. DEFLATE supports ``fixed'' and ``dynamic'' Huffman
codes. Fixed Huffman codes are defined in the DEFLATE specification
and can be simply hard-coded in the compressor/decompressor
software. Dynamic Huffman codes, on the other hand, are created by the
compressor and their information is serialized and stored in the block
header. The compressed data is then run through an implementation of
the LZ77 algorithm from earlier.

@@

The primary means of compression, however, is through model/residual
coding. For each block of raw input, the FLAC compressor tries to fit
the signal to a parameterized function, such as a polynomial for
instance, and then stores the function. It then computes the
residuals, or the differences between the raw audio signal and the
parameterized function at each sample. This residual data has a
particular distribution which is efficiently stored using a specific
form of Huffman codes, called Rice codes. \cite{flac}

>>>>

\usepackage{graphicx}
\usepackage{amssymb, amsmath, amsthm, amsfonts}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{thmtools}
\usepackage{tikz}
\usepackage[pdfusetitle]{hyperref}
\usepackage{cleveref}
\usepackage{apacite}

@@

\title{A Review of Lossless Data Compression Methods}
\author{Jack Stanek}

@@

In the age of ever-expanding quantities of digital data, the ability
to store this data efficiently is more important than ever. Throughout
the history of the field of computer science, algorithms of varying
complexities and efficacies have been created in order to reduce the
amount of space required to store arbitrary strings of binary
data. This wide variety of methods are optimized for different
purposes: some algorithms attempt to reduce the amount of space
required to store raster images; some try to optimize audio storage;
some can optimize genetic data to an extremely high degree; and still
more attempt to work well for general-purpose use. Within the scope of
this paper, I will attempt to give an overview of a handful of popular
compression schemes, including general purpose in addition to
domain-specific methods, as well as a theoretical overview of the
limits and potential benefits of compression in general. This paper
will only cover lossless data encoding: that is, alogrithms which can
reproduce a perfect copy of the input data from its output. Lossy
compression schemes exist which can achieve incredible compression
ratios, but their methods typically include psychological ``tricks''
which allows the algorithm to discard data which is less noticable to
human senses. As this is a mathematics paper and not a psychology
paper, these methods are outside the scope of this document.

@@

Data compression has many benefits as quickly reviewed above; however,
many limitations also exist. Note that for simplicity, we will limit
our discussion of compression schemes to binary schemes, as these are
the most widespread in applications in the real world (i.e.,
compression of computer files.) The first limitation we shall discuss
is the impossibility of a universally applicable, or ``perfect,''
compression scheme.

@@

The Lempel-Ziv compression algorithm, introduced in two variations in
1977 and 1978, is one of the most widespread compression schemes in
use, especially when including its large family of derivative
methods. Algorithms based on the Lempel-Ziv algorithms are relatively
efficient and work well with a wide variety of inputs.

The algorithm is based on a dictionary-style approach; that is to say,
it creates a dictionary of codewords within the input data as the
algorithm progresses. The LZ77 variant uses a ``sliding-window''
technique to construct its dictionary as it compresses. \cite{lz77}
For each set of bytes it attempts to compress, it inspects its sliding
window, which extends some set amount behind the bytes under
consideration. In this way, the algorithm builds a prefix code as it
proceeds through the file. A ``prefix'' can be one or more characters
long. For each point in the input data, the algorithm consults the
prefix table of prefixes contained in the sliding window, which is a
set amount of data prior to the current point. If the current prefix
is found in the prefix table, the algorithm outputs a 3-tuple
indicating how far back in the window the prefix occurred, the length
of the prefix, and finally the character which comes after the
prefix. This means that the very first characters in the input data
will be encoded as 1-character prefixes. (For our very first
characters in the input data, since there can't be any prefixes in the
sliding window yet, the encoded form will consist of zero-length
prefixes followed by the input character literals.)

To illustrate: suppose we are to encode the string ``oodles of
noodles'' using LZ77, with a window of length 8. The string is 16
characters long, requiring 16 bytes to store. The current character is
set to ``o'', the first character of the string. However, since there
are two ``o'' characters in a row, the algorithm can ``overshoot'' the
end of the window in order to efficiently implement run-length
encoding (that is, recording a character and the number of copies
folloiwng it.) Thus, \texttt{(2, 0, 'o')} is emitted, and the current
character is incremented by two. The algorithm then emits
single-character prefix 3-tuples for \texttt{d, l, e, s, ' ',} (the
space character), \texttt{o}, and \texttt{f}. At

The LZ78 algorithm is functionally equivalent to the LZ77 algorithm,
but it forgoes the sliding window altogether, and instead uses a
``linked dictionary''. \cite{lz78} As the compressor processes the
input data, it maintains a matching index to longest prefix already
stored in the dictionary. If the prefix created by appending the
current character to the last matched prefix is not found in the
dictionary, the algorithm stores the character at the next available
index slot along with a pointer to the index of the original prefix,
and outputs the index of the longest matching prefix followed by the
character. Algorithmically, this method is equivalent to LZ77.

@@

LZW is able to achieve this speed by using a single symbol table for
the entire input, adaptively building up as the algorithm processes
it. Instead of using a sliding window, LZW instead begins by
initializing a string table with every single-character prefix. As it
parses the input, it uses a greedy algorithm to match strings from the
input to the strings already seen in the table. It does this by
checking if some prefix in the input has already been seen in the
table; if so, it appends the immediately next character from the input
to the prefix and queries the string table again. If, on the other
hand, the appended prefix is \emph{not} found, the algorithm outputs
the code of the prefix \emph{without} the next character appended (the
code for which is retrieved from the string table) and then adds the
appended prefix to the string table. The current prefix is finally set
to the last character of the prefix and the process repeats.

@@

An exmaple of an efficient implementation of a compression algorithm
can be seen in the DEFLATE format, codified in the RFC 1951
memo. \cite{RFC1951} DEFLATE is not ``standardized'' by any recognized
governing body, though it is very widely used and has become a de
facto standard by virtue of its use in the \texttt{gzip} compression
utility, as well as in the ubiquitous \texttt{zlib} library. RFC 1951
specifies the technical minutia of the implementation of the file
format, as well as its algorithm itself. As such, it is not useful to
describe the full file format within this paper, and so only the
algorithm will be outlined.

Briefly, compression proceeds as follows: for a given block of data
within the input, the algorithm constructs a Huffman prefix coding for
the input alphabet. Information about the prefix coding is then stored
within the output block header, which contains metadata about the
compressed output. DEFLATE supports ``fixed'' and ``dynamic'' Huffman
codes. Fixed Huffman codes are defined in the DEFLATE specification
and can be simply hard-coded in the compressor/decompressor
software. Dynamic Huffman codes, on the other hand, are created by the
compressor and their information is serialized and stored in the block
header. The compressed data is then run through an implementation of
the LZ77 algorithm. As such, this combination of prefix coding with a
powerful compression scheme allows compression tools to squeeze every
last bit of possible compression out of their data.

@@

The primary means of compression, however, is through model/residual
coding. For each block of raw input, the FLAC compressor tries to fit
the signal to a parameterized function, typically a polynomial, and
then stores the function. It then computes the residuals, or the
differences between the raw audio signal and the parameterized
function at each sample. This residual data has a particular
distribution which is efficiently stored using a specific form of
Huffman codes, called Rice codes. \cite{flac}

### 2018-12-08 presentation/presentation.tex
Updates

  A data compression scheme refers to any system for reducing the
  number of symbols required to represent arbitrary information.

  Some commonplace (possibly lossy!) examples:

@@

  Clearly, this list is not exhaustive, but it provides an intuition
  to how these systems function.
\end{frame}

\begin{frame}
  \subsection{Theoretical Background}
  Information can be represented as a series of symbols
  \begin{itemize}
    \item Text is a string of letters in langauge's alphabet
    \item Numbers are strings of Arabic numerals
  \end{itemize}
  For some alphabet $\Sigma$, the set of all compression schemes
  $C_\Sigma$ is the set of bijections
  $c : \Sigma^n \to \bigcup_{k = 1}^{n} \Sigma^k$ for all
  $n \in \mathbb{N}$ such that $n \geq 2$. A compression scheme is
  considered ``perfect'' if and only if its image is the set of all
  strings of length strictly less than that of the domain.
\end{frame}

\begin{frame}
  \subsection{Impossibility of perfect compression}
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

\begin{frame}

\end{frame}

>>>>

  Informally, a data compression scheme is some method for reducing
  the amount of space required to store a given amount of data.

  Some commonplace examples:

@@

  This brief list includes some lossy methods, though this
  presentation will cover only lossless methods.
\end{frame}

\begin{frame}
  \subsection{Theoretical Background}
  Information can be represented as a series of symbols
  \begin{itemize}
  \item Text is a string of letters in langauge's alphabet
  \item Numbers are strings of Arabic numerals
  \end{itemize}
  For some alphabet $\Sigma$, the set of all compression schemes
  $C_\Sigma$ is the set of bijections
  $c : \Sigma^n \to \bigcup_{k = 1}^{n} \Sigma^k$ for all
  $n \in \mathbb{N}$ such that $n \geq 2$.
\end{frame}

\begin{frame}
  \subsection{Impossibility of perfect compression}
  A compression scheme is considered ``perfect'' if and only if its
  image is the set of all strings of length strictly less than that of
  the domain.
  \begin{Theorem}
    No perfect compression scheme exists.
  \end{Theorem}
\end{frame}

\begin{frame}
  \subsection{Maximum Compressibility of Data}
  \begin{itemize}
  \item This result suggests there is a limit to how small compressed
    data can be.
  \item Shannon's Noiseless Coding gives a lower bound.
    \begin{itemize}
    \item The average word size of a prefix code on a channel is
      greater than or equal to its entropy.
    \item Since human data sources are typically very regular
      (i.e. have low information entropy), data compression methods
      are able to greatly reduce the sizes of these sorts of data.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \section{Lempel-Ziv Based Algorithms}
  \subsection{LZ}
  \frametitle{Lempel-Ziv compression algorithms}
  LZ77 and LZ78 (a.k.a. LZ1 and LZ2, a.k.a. LZ)
  \begin{itemize}
  \item Introduced in 1977 and 1978 by Abraham Lempel and Jakob Ziv
  \item Dictionary block compressor family
  \item Uses a sliding-window approach
  \item Produces 3-tuples representing prefixes
  \end{itemize}
\end{frame}

\begin{frame}
  \subsection{LZ77 Example}
  Suppose we are to encode the tongue-twister ``Can you can a can as a
  canner can can a can?'' with a window size of 8.
\end{frame}

\begin{frame}
  \section{FLAC}
  \frametitle{Free Lossless Audio Codec}
  The Free Lossless Audio Codec is a domain-specific compression
  scheme for hi-fi audio.
\end{frame}
